--- 
title: "Bushfire-specific PM~2.5~ V1.3 Dataset (Last edited on `r Sys.Date()`)"
author: "Bushfire Smoke Exposures Team"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(lubridate)
library(terra)
library(data.table)
library(ggplot2)
# displayed date and location
dd <- as.Date("2013-10-17")
place_name <- "Lithgow (NSW)"
place_feature <- "This location was affected by the nearby State Mine bushfire" # for caption
crds <- data.frame(lon = 150.162346, lat = -33.488082)

caption_raster <- sprintf('Raster map of total PM2.5 (ug/m3) for Australia on %s. %s is marked with a white cross, corresponding to the location of the STL decomposition shown below.', dd, place_name)
caption_stl <- sprintf('STL decomposition of total PM2.5 (ug/m3) into trend, seasonal and remainder components for pixel at %s. Dotted red line indicates the date %s, corresponding to the above raster map. %s. (Note the scale for y axis is not fixed.)', place_name, dd, place_feature)

infile <- sprintf("~/CARDAT/ResearchProjects_CAR/Bushfire_Smoke_PM_Exposures/Bushfire_specific_PM25_Aus_2001_2020_v1_3/data_derived/bushfiresmoke_v1_3_%04i_compressed_20231130_7.nc", year(dd))

v_crds <- project(vect(crds, crs = "epsg:4326"), "epsg:3577")
```

# Overview

<style type="text/css">
p.caption {
  color: grey;
  font-size: smaller;
}
</style>

The Bushfire-specific PM~2.5~ V1.3 datasets are a daily raster time-series of estimated PM~2.5~ concentration covering Australia. It additionally includes a decomposition of the estimated PM~2.5~ into seasonal, trend and remainder components, and a series of flags that may be used to predict bushfire smoke events.

```{r img-pm25, fig.cap=caption_raster}
r_plt <- rast(infile, "pm25_pred")
r_plt <- r_plt[[which(time(r_plt) == dd)]]
r_plt <- terra::project(r_plt, "epsg:4326")

plot(r_plt["pm25_pred"])
points(project(v_crds, r_plt), cex = 1.5, col = "white", pch = 4)
```

```{r img-stl, fig.cap=caption_stl}
vars <- c("pm25_pred", "trend", "seasonal", "remainder")

dat <- lapply(vars, function(i){
  r <- rast(infile, i)
  e <- transpose(extract(r, v_crds, ID = F, ))
  setDT(e)
  setnames(e, "value")
  e[, `:=`(date = time(r),
           variable = i)]
  return(e)
})
dat <- rbindlist(dat)
setcolorder(dat, c("variable", "date", "value"))
dat[, variable := factor(variable, levels = c("remainder", "seasonal", "trend", "pm25_pred"))]

ggplot(dat, aes(x = date, y = value)) + 
  geom_line() + 
  facet_wrap(~ variable, ncol = 1, scales = "free") + 
  geom_vline(xintercept = dd, col = "red", lty = "dotted") +
  ylab(substitute("PM~2.5~ (ug/m^3^)"))

```

The initial dataset ([Bushfire_specific_PM25_Aus_2001_2020_v1_3][dset-2001-2020-metadata]) was modelled and predicted on 2001-2020 data covering mainland Australia and Tasmania in a 5km resolution grid, using a random forest model. This PM2.5 prediction was decomposed using the STL method to trend, seasonal and remainder components for each pixel. A series of binary flags were also extracted for each pixel, indicating presence of fires, dust or high temperature in or near the pixel.

Subsequently, this model was used to extend predictions to the years 2020-2023 ([Bushfire_specific_PM25_Aus_2020_2023_v1_3][dset-2020-2023-metadata]) with an accompanying STL decomposition.

The dataset is available on request for research purposes. It is managed by the [Clean Air Research Data and Analysis Technology][cardat-home] (CARDAT) team of the [Centre for Safe Air][csa-home] (CSA). Please submit an access request through the [CARDAT Data Access Form][cardat-form].

## Acknowledgements {-}

The development of the **Bushfire-specific PM~2.5~ 2001-2020 V1.3** dataset and associated **Bushfire-specific PM~2.5~ for locations 2001-2020 V1.3 Workflow** tool was funded by the [Australian Research Data Commons][ardc-home] (ARDC), as part of the [Bushfire Data Challenges - Assessing the impact of bushfire smoke on health project][ardc-project].

We acknowledge the Sydney Informatics Hub and the University of Sydney’s high performance computing cluster Artemis for providing the high performance computing resources on which the **Bushfire-specific PM~2.5~ V1.3** datasets were generated.
 
We would also like to thank Dana Jordan for her assistance during the QA/QC process.

<!-- Links -->
<!-- Dataset pages -->
[dset-2001-2020-doi]: https://doi.org/10.17605/OSF.IO/WQK4T 
[dset-2020-2023-doi]: https://doi.org/10.17605/OSF.IO/47C6B
[dset-2001-2020-metadata]: https://cardat.github.io/data_inventory/bushfire_smoke_pm_exposures.html#bushfire_specific_pm25_aus_2001_2020_v1_3
[dset-2020-2023-metadata]: https://cardat.github.io/data_inventory/bushfire_smoke_pm_exposures.html#bushfire_specific_pm25_aus_2020_2023_v1_3

<!-- Organisation pages -->
[csa-home]: https://safeair.org.au/
[cardat-home]: https://cardat.github.io/
[cardat-form]: https://cardat.github.io/data_access_request_form.html
[ardc-home]: https://ardc.edu.au/
[ardc-project]: https://doi.org/10.47486/DC008

<!-- Other links -->
[paper-dset]: https://doi.org/10.3390/atmos15111341
[paper-exceptional-events]: 10.1016/j.heliyon.2024.e24532