--- 
title: "Bushfire-specific PM~2.5~ V1.3 data (Last edited on `r Sys.Date()`)"
author: "Bushfire Smoke Exposures Team"
css: css/style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.path = 'main_images/')
library(lubridate)
library(terra)
library(data.table)
library(ggplot2)
library(gganimate)

# displayed date and location for home page
dd <- as.Date("2013-10-18")
place_name <- "Lithgow (NSW)"
place_feature <- "This location was affected by the nearby State Mine bushfire" # for caption
crds <- data.frame(lon = 150.162346, lat = -33.488082)

caption_raster <- sprintf('Raster map of total PM~2.5~ (µg/m^3^) for Australia on %s. %s is marked with a red diamond, corresponding to the location of the STL decomposition shown below.', dd, place_name)
caption_stl <- sprintf('STL decomposition of total PM~2.5~ (µg/m^3^) into trend, seasonal and remainder components for the pixel at %s. Dotted red line indicates the date %s, corresponding to the above raster map. %s. (Note the scale for y axis is not fixed.)', place_name, dd, place_feature)

infile <- file.path(datadir,
                    sprintf("Bushfire_specific_PM25_Aus_2001_2020_v1_3/data_derived/bushfiresmoke_v1_3_%04i_compressed_20231130_7.nc", 
                            year(dd))
)

v_crds.wgs84 <- vect(crds, crs = "epsg:4326")
v_crds <- project(v_crds.wgs84, "epsg:3577")
```

# Overview

The Bushfire-specific PM~2.5~ V1.3 datasets are a set of daily time-series rasters of estimated total PM~2.5~ concentration covering Australia (Fig. \@ref(fig:img-pm25)). These additionally include a seasonal-trend decomposition using LOESS (STL) of the total PM~2.5~ into seasonal, trend and remainder components (Figure \@ref(fig:img-stl)), as well as a series of binary flags. Both the decomposed PM~2.5~ components and flags may be used in identification and quantification of bushfire-related PM~2.5~ exposures. The original dataset accompanies the following publication:

> Borchers-Arriagada, N., Morgan, G.G., Buskirk, J.V., Gopi, K., Yuen, C., Johnston, F.H., Guo, Y., Cope, M. and Hanigan, I.C. (2024) ‘Daily PM2.5 and Seasonal-Trend Decomposition to Identify Extreme Air Pollution Events from 2001 to 2020 for Continental Australia Using a Random Forest Model’, *Atmosphere*, *15*(1341). Available at: <https://doi.org/10.3390/atmos15111341>.

```{r img-pm25, fig.cap=caption_raster, cache = TRUE}
r_plt <- rast(infile, "pm25_pred")
r_plt <- r_plt[[which(time(r_plt) == dd)]]
r_plt <- terra::project(r_plt, "epsg:4326")

focus_bounds <- buffer(v_crds.wgs84, 500000)
r_plt.focus <- crop(r_plt, focus_bounds)


plot(r_plt.focus, 
     plg=list(title = expression("PM"[2.5]*" (µg/m"^3*")")))
points(v_crds.wgs84, cex = 1, col = "red", pch = 18)
sbar(100, type = "bar", xy = c(153.9, -29.6), below = "km")
inset(r_plt, box=ext(focus_bounds), scale=.4, loc="bottomright", background="grey90", pbox=list(lwd=1, lty=1, col="red"))
```

```{r img-stl, fig.cap=caption_stl, cache = TRUE}
vars <- c("pm25_pred", "trend", "seasonal", "remainder")

dat.index <- lapply(vars, function(x){
  r <- rast(infile, x)                 # read raster variable
  e <- extract(r, project(v_crds, r), ID = F)  # extract at given locations
  setDT(e)
  e[, name := place_name]
  e <- melt(e, id.vars = "name", variable.factor = F)    # reshape to long
  e$date <- time(r)               # add date column
  e$variable <- x                 # overwrite variable column with sensible name
  return(e)
})
dat.index <- rbindlist(dat.index)
setcolorder(dat.index, c("variable", "date", "value"))
dat.index[, variable := factor(variable, levels = c("remainder", "seasonal", "trend", "pm25_pred"),
                         labels = c("Remainder", "Seasonal", "Trend", "Total PM2.5"))]

ggplot(dat.index, aes(x = date, y = value)) + 
  geom_line() + 
  facet_wrap(~ variable, ncol = 1, scales = "free") + 
  geom_vline(xintercept = dd, col = "red", lty = "dotted") +
  xlab("Date") +
  ylab(substitute("PM"[2.5]*" (µg/m"^3*")"))

```

Read more about the dataset and its development in the [Metadata](#metadata) and [Methodology](#methodology).

The dataset is available on request for research purposes. It is managed by the [Clean Air Research Data and Analysis Technology][cardat-home] (CARDAT) team of the [Centre for Safe Air][csa-home] (CSA). Please read the full [data sharing policy](#dataset-licencing) and submit an access request through the [CARDAT Data Access Form][cardat-form].

## Acknowledgements {-}

The development of the *Bushfire-specific PM~2.5~ 2001-2020 V1.3* dataset was funded by the [Australian Research Data Commons][ardc-home] (ARDC), as part of the [Bushfire Data Challenges - Assessing the impact of bushfire smoke on health project][ardc-project].

We acknowledge the Sydney Informatics Hub and the University of Sydney’s high performance computing cluster Artemis for providing the high performance computing resources on which the *Bushfire-specific PM~2.5~ V1.3* datasets were generated.
 
We would also like to thank Dana Jordan for her assistance during the QA/QC process.

<!-- Links -->
<!-- Dataset pages -->
[dset-2001-2020-doi]: https://doi.org/10.17605/OSF.IO/WQK4T 
[dset-2020-2023-doi]: https://doi.org/10.17605/OSF.IO/47C6B
[dset-2001-2020-metadata]: https://cardat.github.io/data_inventory/bushfire_smoke_pm_exposures.html#bushfire_specific_pm25_aus_2001_2020_v1_3
[dset-2020-2023-metadata]: https://cardat.github.io/data_inventory/bushfire_smoke_pm_exposures.html#bushfire_specific_pm25_aus_2020_2023_v1_3

<!-- Organisation pages -->
[csa-home]: https://safeair.org.au/
[cardat-home]: https://cardat.github.io/
[cardat-form]: https://cardat.github.io/data_access_request_form.html
[ardc-home]: https://ardc.edu.au/
[ardc-project]: https://doi.org/10.47486/DC008

<!-- Other links -->
[paper-dset]: https://doi.org/10.3390/atmos15111341
[paper-exceptional-events]: https://doi.org/10.1016/j.heliyon.2024.e24532